Massively Parallel Monte Carlo Methods for Discrete Linear and Nonlinear Systems

### 2012-11-03 doc/prelim/presentation/prelim_presentation.tex
adding some references to the presentation
combined Matrix-Free and FAD slides

  \pause
  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
    {Predictive nuclear reactor analysis enables...}
    \begin{itemize}
    \item Tighter design tolerance for improved thermal performance
      and efficiency
    \item Higher fuel burn-up
    \item High confidence in accident scenario models
    \end{itemize}
  \end{beamerboxesrounded}

  \pause
  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
    {Multiple physics simulations are complicated...}
    \begin{itemize}
    \item Neutronics, thermal hydraulics, computational fluid
      dynamics, structural mechanics, and many other physics
    \item Consistent models yield nonlinearities in the variables
      through feedback effects
    \item Tremendous computational resources are required with
      $O(\sn{1}{9})$ element meshes and $O(100,000)+$ cores used in
      today's simulations.
    \end{itemize}
  \end{beamerboxesrounded}

@@

    \pause

@@

  \pause
  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
    {A Requirement} Assert that $\ve{A}$ is \textit{nonsingular}. The
    solution is then:
    \[
    \ve{x} = \ve{A}^{-1}\ve{b}
    \]
  \end{beamerboxesrounded}

@@

  \begin{itemize}
  \item Powerful class of iterative methods
  \item Provides theory that encapsulates most other iterative methods
  \item Leveraged in many modern physics codes at the petascale
  \end{itemize}

  \pause
  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
    {Search Subspace $\mathcal{K}$} Extract the solution from the
    search subspace:
    \[
    \tilde{\ve{x}} = \ve{x}_0 +
    \boldsymbol{\delta},\ \boldsymbol{\delta} \in \mathcal{K}
    \]
  \end{beamerboxesrounded}

  \pause
  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
    {Constraint Subspace $\mathcal{L}$} Constrain the extraction with
    the constraint subspace by asserting orthogonality with the
    residual:
    \[
    \langle \tilde{\ve{r}},\ve{w} \rangle = 0,\ \forall \ve{w} \in
    \mathcal{L}
    \]  
  \end{beamerboxesrounded}

@@

  \pause
  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
    {Minimization Property}

@@

  \pause
  \begin{figure}[htpb!]
    \begin{center}
      \scalebox{0.7}{ \input{matvec_proc_1.pdftex_t} }
    \end{center}
    \caption{\textbf{Components of multiply operation owned by process
        1.} }
    \label{fig:matvec_proc_1}
  \end{figure}

@@

  \begin{itemize}
  \item Widely used in practice
  \item Krylov methods require only the action of the operator
  \item Parallelism achieved through a handful of operations
  \item Global reduction operations observed not to impede scalability
    \begin{itemize}
    \item Dot product
    \item Vector norms
    \end{itemize}
  \item Nearest neighbor computations have poor algorithmic strong
    scaling
    \begin{itemize}
    \item Matrix-vector multiply
    \item Weak scaling is better
    \end{itemize}
  \item Short and long recurrence relations available for
    orthogonalization
  \end{itemize}

@@

  \begin{itemize}
  \item Neumann-Ulam methods bound by the Central Limit Theorem
  \item Halton proposed an iterative residual method in 1962
  \item Iteration error decoupled from Monte Carlo error
  \item Exponential convergence
  \end{itemize}

@@

    \begin{itemize}
    \item Looked at 4 communication patterns:
      \begin{itemize}
      \item Fully-locking synchronous
      \item asynchronous-send/synchronous-receive
      \item master/slave
      \item binary tree master/slave
      \end{itemize}
    \item Binary tree master/slave performed best but race conditions observed
    \item A later improvement to their work showed a fully asynchronous
      pattern to work best
    \item Poor scaling for unbalanced problems
    \end{itemize}

@@

  \pause
  \begin{itemize}
  \item Many multiphysics problems of interest are nonlinear
  \item Segregated methods lack the consistency of fully implicit
    methods
  \item Newton methods often leverage Krylov solvers
    \begin{itemize}
    \item Robust implementations
    \item No operator required
    \item Memory intensive
    \end{itemize}
  \item Monte Carlo methods need the full operator
  \item Automatic construction of the linear operator available
    \begin{itemize}
    \item Ideal for Monte Carlo
    \item Relaxes memory requirements
    \item Potential scaling improvements
    \item Resiliency benefits
    \end{itemize}
  \end{itemize}

@@

  \begin{itemize}
  \item A form of inexact Newton methods
  \end{itemize}

  \[
  ||\ve{J}(\ve{u}^k)\delta \ve{u}^k + \ve{F}(\ve{u}^k)|| \leq \eta^k
  ||\ve{F}(\ve{u}^k)||,\ \eta^k \in [0,1)
    \]

    \begin{itemize}
    \item Choose a Krylov method to solve for the Newton correction
    \item GMRES with a long recurrence relation observed as more
      robust
    \item Generates a monotonically decreasing residual from
      maintaining the optimization and orthogonality conditions
    \end{itemize}

    \pause Where does the Jacobian come from?  \pause
    \begin{itemize}
    \item The Jacobian can come from hand-coded derivatives
      \begin{itemize}
      \item Tedious and error prone
      \item Repeated for each equation set and hard to do for
        multiphysics
      \end{itemize}
    \end{itemize}

@@

%%---------------------------------------------------------------------------%%
\begin{frame}{Matrix-Free Approximation}

  \begin{itemize}
  \item Krylov methods only need the action of the linear operator
  \item The action of the Jacobian can be approximated using a forward
    difference
  \end{itemize}

@@

  \begin{itemize}
  \item Forms the basis of Jacobian-Free Newton-Krylov (JFNK) methods
    \begin{itemize}
    \item Sensitive to scaling and discretization error
    \item Still have to form part of Jacobian periodically for
      preconditioning
    \item Eventually break even with generating and storing the full
      Jacobian
    \end{itemize}
  \end{itemize}

%%---------------------------------------------------------------------------%%
\begin{frame}{Automatic Differentiation}

  \begin{itemize}
  \item Automatically generate Jacobians from nonlinear function
    evaluations
    \begin{itemize}
    \item Overload math operators and apply the chain rule (FAD)
    \item Yields evaluations as accurate as function discretization
    \end{itemize}
  \item Modern packages take an element-level assembly approach
    \begin{itemize}
    \item Function evaluations are local, communication builds global
      data
    \end{itemize}
  \end{itemize}

  \[
  \ve{J}(\ve{u}) = \sum_{i=1}^N \ve{Q}^T_i \ve{J}_k \ve{P}_i
  \]
  \[
  e_k : \mathbb{R}^{n_k} \rightarrow \mathbb{R}^{m_k},\ \ve{J}_{k_i} =
  \partial e_{k_i} / \partial P_i u,\ \ve{P} \in \mathbb{R}^{m_{k_i}
    \times N}
  \]

  \begin{itemize}
  \item Performance studies give acceptable results for use in
    large-scale, production physics codes
  \end{itemize}

  \pause
  \begin{algorithm}[H]
    \begin{algorithmic}[1]
      \STATE $k := 0$ 
      \WHILE{$||\ve{F}(\ve{u}^{k})|| > \epsilon
        ||\ve{F}(\ve{u}^{0})||$} 
      \STATE $\ve{J}(\ve{u}^{k}) \leftarrow AD(\ve{F}(\ve{u}^k))$ 
      \COMMENT{Automatic differentiation} 
      \STATE $\ve{J}(\ve{u}^k) \delta \ve{u}^k = -\ve{F}(\ve{u}^{k})$
      \COMMENT{Solve for the Newton correction with MCSA} 
      \STATE $\ve{u}^{k+1} \leftarrow \ve{u}^k + \delta \ve{u}^k$ 
      \STATE $k \leftarrow k+1$ 
      \ENDWHILE
    \end{algorithmic}
    \caption{FANM}
  \end{algorithm}

  \pause
  \begin{itemize}
  \item Robustness of Newton's method (inexact)
  \item Accuracy and convenience of FAD
  \item Parallelism, memory, and resiliency benefits of MCSA
  \item Requires only nonlinear function evaluations
  \end{itemize}

@@

  \pause \small{ \[
    \begin{array}{lccccccccccccccc}
      \ve{A}\ values: & 2 & 8 & 4 & 5 & 1 & 2 & 1 & 1 & 3 & 7 & 2 & 4
      & 9 & 9 & 1 \\ column: & 1 & 3 & 1 & 2 & 4 & 2 & 3 & 5 & 3 & 4 &
      6 & 4 & 5 & 5 & 6 \\ row\ start: & 1 & 3 & 6 & 9 & 12 & 14 & 16
      \\
    \end{array}
    \]}

@@

>>>>

  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
    {Predictive nuclear reactor analysis enables...}
    \begin{itemize}
    \item Tighter design tolerance for improved thermal performance
      and efficiency
    \item Higher fuel burn-up
    \item High confidence in accident scenario models
    \end{itemize}
  \end{beamerboxesrounded}

  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
    {Multiple physics simulations are complicated...}
    \begin{itemize}
    \item Neutronics, thermal hydraulics, computational fluid
      dynamics, structural mechanics, and many other physics
    \item Consistent models yield nonlinearities in the variables
      through feedback effects
    \item Tremendous computational resources are required with
      $O(\sn{1}{9})$ element meshes and $O(100,000)+$ cores used in
      today's simulations.
    \end{itemize}
  \end{beamerboxesrounded}

@@

@@

  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
    {A Requirement} Assert that $\ve{A}$ is \textit{nonsingular}. The
    solution is then:
    \[
    \ve{x} = \ve{A}^{-1}\ve{b}
    \]
  \end{beamerboxesrounded}

@@

  \begin{itemize}
  \item Powerful class of iterative methods (Saad,2003)
  \item Provides theory that encapsulates most other iterative methods
  \item Leveraged in many modern physics codes at the petascale
  \end{itemize}

  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
    {Search Subspace $\mathcal{K}$} Extract the solution from the
    search subspace:
    \[
    \tilde{\ve{x}} = \ve{x}_0 +
    \boldsymbol{\delta},\ \boldsymbol{\delta} \in \mathcal{K}
    \]
  \end{beamerboxesrounded}

  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
    {Constraint Subspace $\mathcal{L}$} Constrain the extraction with
    the constraint subspace by asserting orthogonality with the
    residual:
    \[
    \langle \tilde{\ve{r}},\ve{w} \rangle = 0,\ \forall \ve{w} \in
    \mathcal{L}
    \]  
  \end{beamerboxesrounded}

@@

  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
    {Minimization Property}

@@

  \begin{figure}[htpb!]
    \begin{center}
      \scalebox{0.7}{ \input{matvec_proc_1.pdftex_t} }
    \end{center}
    \caption{\textbf{Components of multiply operation owned by process
        1.} }
    \label{fig:matvec_proc_1}
  \end{figure}

@@

  \begin{itemize}
  \item Widely used in practice
  \item Krylov methods require only the action of the operator
  \item Parallelism achieved through a handful of operations
  \item Global reduction operations observed not to impede scalability
    (Gropp,2001)
    \begin{itemize}
    \item Dot product
    \item Vector norms
    \end{itemize}
  \item Nearest neighbor computations have poor algorithmic strong
    scaling
    \begin{itemize}
    \item Matrix-vector multiply
    \item Weak scaling is better
    \end{itemize}
  \item Short and long recurrence relations available for
    orthogonalization
  \end{itemize}

@@

  \begin{itemize}
  \item Neumann-Ulam methods bound by the Central Limit Theorem
  \item Halton proposed an iterative residual method in 1962
  \item Iteration error decoupled from Monte Carlo error
  \item Exponential convergence (Evans,2003)
  \end{itemize}

@@

    \begin{itemize}
    \item Looked at 4 communication patterns:
      \begin{itemize}
      \item Fully-locking synchronous
      \item asynchronous-send/synchronous-receive
      \item master/slave
      \item binary tree master/slave
      \end{itemize}
    \item Binary tree master/slave performed best but race conditions observed
    \item A later improvement to their work showed a fully asynchronous
      pattern to work best
    \item Poor scaling for unbalanced problems for all schemes
    \end{itemize}

@@

  \begin{itemize}
  \item Many multiphysics problems of interest are nonlinear
  \item Segregated methods lack the consistency of fully implicit
    methods
  \item Newton methods often leverage Krylov solvers
    \begin{itemize}
    \item Robust implementations
    \item No operator required
    \item Memory intensive
    \end{itemize}
  \item Monte Carlo methods need the full operator
  \item Automatic construction of the linear operator available
    \begin{itemize}
    \item Ideal for Monte Carlo
    \item Relaxes memory requirements
    \item Potential scaling improvements
    \item Resiliency benefits
    \end{itemize}
  \end{itemize}

@@

  \begin{itemize}
  \item Choose a Krylov method to solve for the Newton correction
  \item GMRES with a long recurrence relation observed as more
    robust (Knoll,2004)
  \item Generates a monotonically decreasing residual from
    maintaining the optimization and orthogonality conditions
  \end{itemize}

  \pause Where does the Jacobian come from?
  \begin{itemize}
  \item The Jacobian can come from hand-coded derivatives
    \begin{itemize}
    \item Tedious and error prone
    \item Repeated for each equation set and hard to do for
      multiphysics
    \end{itemize}
  \end{itemize}

@@

%%---------------------------------------------------------------------------%%
\begin{frame}{Jacobian Approximations}

  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
    {Matrix-Free Approximation}
    \begin{itemize}
    \item Krylov methods only need the action of the linear operator
    \end{itemize}

@@

    \begin{itemize}
    \item Forms the basis of Jacobian-Free Newton-Krylov (JFNK) methods
      \begin{itemize}
      \item Sensitive to scaling and discretization error (Kelly,1995)
      \item Eventually break even with generating and storing the full
        Jacobian (Knoll,1994)
      \end{itemize}
    \end{itemize}
  \end{beamerboxesrounded}

  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
    {Automatic Differentiation}
    \begin{itemize}
    \item Automatically generate Jacobians from nonlinear function
      evaluations
      \begin{itemize}
      \item Overload math operators and apply the chain rule (FAD)
      \item Yields evaluations as accurate as function discretization
      \end{itemize}
    \item Performance studies give acceptable results for use in
      large-scale, production physics codes (Bartlett,2006)
    \end{itemize}
  \end{beamerboxesrounded}

@@

  \small{ \[
    \begin{array}{lccccccccccccccc}
      \ve{A}\ values: & 2 & 8 & 4 & 5 & 1 & 2 & 1 & 1 & 3 & 7 & 2 & 4
      & 9 & 9 & 1 \\ column: & 1 & 3 & 1 & 2 & 4 & 2 & 3 & 5 & 3 & 4 &
      6 & 4 & 5 & 5 & 6 \\ row\ start: & 1 & 3 & 6 & 9 & 12 & 14 & 16
      \\
    \end{array}
    \]}

@@

  \begin{itemize}
  \item Robustness of Newton's method (inexact)
  \item Accuracy and convenience of FAD
  \item Parallelism, memory, and resiliency benefits of MCSA
  \item Requires only nonlinear function evaluations
  \end{itemize}
